Compositional Equivalences Based on Open pNets

### 2022-10-19 2016-Weak-FH-Bisim/WeakBisim.tex
letter and blue

Bisimulation over a symbolic and open model like open pNets or open automata is  different from the classical notion of bisimulation because it cannot rely on the equality over a finite set of action labels. Classical bisimulations require to exhibit, for each transition of one system, a transition of the other system that simulates it. Instead, bisimulation for open automata  relies on the simulation of each open transition of one automaton by \emph{a set of} open transitions of the other one, that should cover all the 
cases where the original transition can be triggered. %\TODO{Eric: is this like in symbolic bisimulation ?}
\ERIC{This is similar to the early and late symbolic bisimulation equivalences for value-passing CCS \cite{HENNESSY:1995}, though we use more general definitions in our setting.}

Compositionality of bisimilarity in our model comes from the specification of the interactions, including actions of the holes. 
\ERIC{This is quite different from the works on contextual equivalences, e.g. \cite{Larsen87, Larsen:91}; we will give more details on this in section \ref{sec:RW}.}
In pNets, synchronisation vectors define the possible interactions between the pNet that fills the hole and the surrounding pNets. In open automata, this is reflected by symbolic hypotheses that depend on the actions of the holes. This additional specification is the price to pay to obtain the compositionality of bisimilarity that cannot be guaranteed in traditional process algebras.

@@

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions of several LTSs with
action of some holes, the transition occurs if some predicates hold, and can involve a 
set of state modifications. This section defines open automata and a bisimulation theory for them. This section is an improved version of the formalism described in \cite{henrio:Forte2016}, extending the automata with a notion of global variable, which makes the state of the automaton more explicit. We also adopt a semantics and logical interpretation of the automata that intuitively can be stated as follows: ``if a transition belongs to an open automaton, any refinement of this transition also belongs to the automaton''. \LUDO{Our open automata are clearly inspired by the work  of De Simone on formatting of SOS rules~\cite{deSimone85}. A precise comparison with related works can be found in Section~\ref{sec:RW}.}

@@

We define now a bisimulation relation tailored to open automata and their parametric nature. This relation relates states of the open automata and guarantees that the related states are observationally equivalent, i.e. equivalent states can trigger transitions with identical action labels. Its key characteristics are 1) the introduction of predicates in the bisimulation relation: the relation between states may depend on the value of the variables; 2)  bisimulation  relates elements of the open transitions and takes into account predicates over variables, actions of the holes, and state modifications.
 We name it FH-bisimulation,
 as a short cut for the ``Formal Hypotheses'' over the holes behaviour manipulated in the
 transitions, but also as a reference to the work of De Simone~\cite{deSimone85},
 that pioneered this idea.
\ERIC{Indeed, our definition uses both hypotheses on the behaviour of holes, as in \cite{deSimone85}, and symbolic manipulation of action expressions, as in symbolic bisimulations of \cite{HENNESSY:1995}. }
%\TODO{Eric: combination of RdS, symb bisim, and more}

@@

\begin{example}
The simulation of one transition by many others is one non-standard aspect of this definition. This is made necessary by the parameterised nature of our model. Consider the following open transition.
\[ \openrule
         {
           \set{\beta}, \True,\subst{y\leftarrow x}}
          {s_1 \OTarrow {\alpha(x)} s_1'}\]
Bisimulation should allow it to be matched by the two following ones (depending on the value of $x$), to prove the bisimulation $\mathcal{R}=\{(s_1,s_2,\True),(s_1',s_2',\True)\}$
\[ \openrule
         {
           \set{\beta}, x\geq 0,\subst{y\leftarrow x}}
          {s_2 \OTarrow {\alpha(x)} s_2'}
\qquad
\openrule
         {
           \set{\beta}, x<0,\subst{y\leftarrow x}}
          {s_2 \OTarrow {\alpha(x)} s_2'}
\]
This example illustrates the necessity of multiple transitions in the definition of bisimulation in a naive and minimalistic way.  It can easily be extended into a non-trivial example with more states and different usage of the variables.
\end{example}

@@

In the same vein as  context systems \cite{Larsen:91}, pNets is a formalism for modular and possibly incomplete  description of concurrent systems. The two formalisms are however different as the theory of contexts relies on a form of rewrite rules, while  pNets rely on parametric automata to express the system behaviour.  
PNets have similar features as context systems \cite{Larsen:91} and static constructs \cite{HuttelL89}. Indeed all these approaches allow for modular and possibly incomplete  description   and  structural composition of systems. The main originality of pNets  compared to these other compositional approaches is the parameterised nature of the specification, which enables reasoning on value-passing systems but also on rich synchronisations that depend on the value of parameters.

@@

\LUDO{
pNets support the refinement checking at the automaton level through a simulation approach, with symbolic evaluation of
the guards and transitions. 
 The definition of simulation on open automata should be  stronger than a classical simulation since it matches a transition with a family of transitions. Such a relation should be able to check the refinement by taking into account state duplication,  transition removal,  guard strengthening,  variable modification. Additionally, composition of 
pNets gives the possibility to either add new holes to a system or  fill holes. A useful simulation relation  should thus support the  comparison of automata that do not have the same number of holes. Designing such a simulation relation is a non-trivial extension that we leave for future work.}

>>>>

Bisimulation over a symbolic and open model like open pNets or open automata is  different from the classical notion of bisimulation because it cannot rely on the equality over a finite set of action labels. Classical bisimulations require to exhibit, for each transition of one system, a transition of the other system that simulates it. Instead, bisimulation for open automata  relies on the simulation of each open transition of one automaton by \emph{a set of} open transitions of the other one, that should cover all the 
cases where the original transition can be triggered. %\TODO{Eric: is this like in symbolic bisimulation ?}
This is similar to the early and late symbolic bisimulation equivalences for value-passing CCS \cite{HENNESSY:1995}, though we use more general definitions in our setting.

Compositionality of bisimilarity in our model comes from the specification of the interactions, including actions of the holes. 
This is quite different from the works on contextual equivalences, e.g. \cite{Larsen87, Larsen:91}; we will provide a detailed comparison in Section~\ref{sec:RW}.
In pNets, synchronisation vectors define the possible interactions between the pNet that fills the hole and the surrounding pNets. In open automata, this is reflected by symbolic hypotheses that depend on the actions of the holes. This additional specification is the price to pay to obtain the compositionality of bisimilarity that cannot be guaranteed in traditional process algebras.

@@

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions of several LTSs with
action of some holes, the transition occurs if some predicates hold, and can involve a 
set of state modifications. This section defines open automata and a bisimulation theory for them. This section is an improved version of the formalism described in \cite{henrio:Forte2016}, extending the automata with a notion of global variable, which makes the state of the automaton more explicit. We also adopt a semantics and logical interpretation of the automata that intuitively can be stated as follows: ``if a transition belongs to an open automaton, any refinement of this transition also belongs to the automaton''. {Our open automata are clearly inspired by the work  of De Simone on formatting of SOS rules~\cite{deSimone85}. A precise comparison with related works can be found in Section~\ref{sec:RW}.}

@@

We define now a bisimulation relation tailored to open automata and their parametric nature. This relation relates states of the open automata and guarantees that the related states are observationally equivalent, i.e. equivalent states can trigger transitions with identical action labels. Its key characteristics are 1) the introduction of predicates in the bisimulation relation: the relation between states may depend on the value of the variables; 2)  bisimulation  relates elements of the open transitions and takes into account predicates over variables, actions of the holes, and state modifications.
 We name it FH-bisimulation,
 as a short cut for the ``Formal Hypotheses'' over the holes behaviour manipulated in the
 transitions, but also as a reference to the work of De Simone~\cite{deSimone85},
 that pioneered this idea.
Indeed, our definition uses both hypotheses on the behaviour of holes, as in \cite{deSimone85}, and symbolic manipulation of action expressions, as in symbolic bisimulations of \cite{HENNESSY:1995}. 
%\TODO{Eric: combination of RdS, symb bisim, and more}

@@

\begin{example}
The simulation of one transition by many others is one non-standard aspect of this definition. This is made necessary by the parameterised nature of our model. Consider the following open transition.
\[ \openrule
         {
           \set{\beta}, \True,\subst{y\leftarrow x}}
          {s_1 \OTarrow {\alpha(x)} s_1'}\]
Bisimulation should allow it to be matched by the two following ones (depending on the value of $x$), to prove that the relation $\mathcal{R}=\{(s_1,s_2,\True),(s_1',s_2',\True)\}$ is a bisimulation.
\[ \openrule
         {
           \set{\beta}, x\geq 0,\subst{y\leftarrow x}}
          {s_2 \OTarrow {\alpha(x)} s_2'}
\qquad
\openrule
         {
           \set{\beta}, x<0,\subst{y\leftarrow x}}
          {s_2 \OTarrow {\alpha(x)} s_2'}
\]
This example illustrates the necessity of multiple transitions in the definition of bisimulation in a naive and minimalistic way.  It can easily be extended into a non-trivial example with more states and different usage of the variables.
\end{example}

@@

In the same vein as  context systems \cite{Larsen:91}, pNets is a formalism for modular and possibly incomplete  description of concurrent systems. The two formalisms are however different as the theory of contexts relies on a form of rewrite rules, while  pNets rely on parametric automata to express the system behaviour.  
pNets have similar features as context systems \cite{Larsen:91} and static constructs \cite{HuttelL89}. Indeed all these approaches allow for modular and possibly incomplete  description   and  structural composition of systems. The main originality of pNets  compared to these other compositional approaches is the parameterised nature of the specification, which enables reasoning on value-passing systems but also on rich synchronisations that depend on the value of parameters.

@@

The pNets formalism supports the refinement checking at the automaton level through a simulation, with symbolic evaluation of
 guards and transitions. 
 The definition of simulation on open automata should be  stronger than a classical simulation since it matches a transition with a family of transitions. Such a relation should be able to check the refinement by taking into account state duplication,  transition removal,  guard strengthening,  variable modification. Additionally, composition of 
pNets gives the possibility to either add new holes to a system or  fill holes. A useful simulation relation  should thus support the  comparison of automata that do not have the same number of holes. Designing such a simulation relation is a non-trivial extension that we leave for future work.
